---
title: "Untitled"
description: |
  A new article created using the Distill format.
author:
  - name: Nora Jones 
    url: https://example.com/norajones
    affiliation: Spacely Sprockets
    affiliation_url: https://example.com/spacelysprokets
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

The United States has a pedestrian fatality crisis. After years of progress, the number of people killed while walking started rising again around 2010, contrary to what's happening outside of the US. A recent article in the New York Times took a closer look at the fatality data and made a convincing case that the crisis specifically is a crisis happening at night. This chart shows the peak of pedestrian fatalities shifting over the course of the year, presumably corresponding with when the sun rises and sets. 

add chart from article

It occurred to me that we can explicitly look at sunrise and sunset times and how they correlate with the time of a crash. This turned out to be a more complicated process than expected and I document the process below.

# Download data
Complete FARS is available as SAS and csv files for download. The files are large and downloads regularly time out. The structure of the folders and files also has some slight variations between years and therefore reading in the files also requires some effort and multiple functions. 

```{r}
library(tidyverse)
library(haven)
library(sf)
library(lutz)
library(suncalc)
library(extrafont)
library(hrbrthemes)
options(timeout=300) #files are large...
```

```{r}
get_fars <- function(year){
  fars_url <- paste0("https://static.nhtsa.gov/nhtsa/downloads/FARS/",
                     year,
                     "/National/FARS",
                     year,
                     "NationalSAS.zip")
  temp_file <- tempfile()
  download.file(fars_url, destfile = temp_file)
  unzip(temp_file, exdir = paste0("data/", year), overwrite = TRUE)
}

walk(2002:2007, safely(get_fars)) #some years will time out and you'll have to try again

```

Now that we have the data downloaded, we read in two tables, the `PERSON` and the `ACCIDENT` files. Unfortunately the structure of those files varies over time and so joining them isn't that straightforward. Additionally, the variables and format for the latitude and longitude are also inconsistent between years.

```{r}
# folder structure is inconsistent between years and we need two functions
# to read in the data
read_FARS_subfolder <- function(year, table_type){
  read_sas(paste0("data/", year, "/FARS", year, "NationalSAS/",table_type, ".sas7bdat"))
}

read_FARS_no_subfolder <-  function(year, table_type){
  read_sas(paste0("data/", year, "/", table_type, ".sas7bdat"))
}

no_subfolder_years <- c(2000:2005, 2016:2019)
subfolder_years <- c(2006:2015, 2020:2021)

accident_00 <- read_FARS_no_subfolder(2000, "accident")
```
The variable in which the latitude and longitude are being saved and their format varies between years. This requires cleanup.

```{r clean-up-2000}
accident_00 <- accident_00 |> 
  mutate(LATITUDE = na_if(LATITUDE, "88888888"), #remove missing
         LATITUDE = na_if(LATITUDE, "99999999"), #remove missing
         LATITUDE = paste0("-",
                           str_sub(LATITUDE, start = 0, end = 2),
                           ".",
                           str_sub(LATITUDE, start = 3)),
         LATITUDE = na_if(LATITUDE, "-NA.NA")) |> 
  mutate(LONGITUD = na_if(LONGITUD, "888888888"),
         LONGITUD = na_if(LONGITUD, "999999999"),
         LONGITUD = paste0(str_sub(LONGITUD, start = 0, end = 3),
                           ".",
                           str_sub(LONGITUD, start = 4)),
         LONGITUD = na_if(LONGITUD, "-NA.NA")) |> 
  mutate(across(c(LATITUDE, LONGITUD), ~ as.numeric(.x)))
```

```{r}
accident_01_05 <- map2_dfr(2001:2005, 
                                  "accident", 
                                  read_FARS_no_subfolder)

accident_01_05 <- accident_01_05 |> 
  mutate(LATITUDE = latitude,
         LONGITUD = longitud)
```

```{r}
accident_06_07 <- map2_dfr(2006:2007, 
                               "accident", 
                               read_FARS_subfolder)
accident_06_07 <- accident_06_07 |> 
  mutate(LATITUDE = latitude,
         LONGITUD = longitud)

accident_08_15 <- map2_dfr(2008:2015, 
                               "accident", 
                               read_FARS_subfolder)

accident_08_15 <- accident_08_15 |> 
  mutate(LATITUDE = if_else(LATITUDE %in% c(99.9999, 88.8888, 77.7777), NA, LATITUDE),
         LONGITUD = if_else(LONGITUD %in% c(999.9999, 888.8888, 777.7777), NA, LONGITUD)) 

accident_16_19 <- map2_dfr(2016:2019, 
                                  "accident", 
                                  read_FARS_no_subfolder)
accident_16_19 <- accident_16_19 |> 
  mutate(LATITUDE = if_else(LATITUDE %in% c(99.9999, 88.8888, 77.7777), NA, LATITUDE),
         LONGITUD = if_else(LONGITUD %in% c(999.9999, 888.8888, 777.7777), NA, LONGITUD))

accident_20_21 <- map2_dfr(2020:2021, 
                               "accident", 
                               read_FARS_subfolder)

accident_20_21 <- accident_20_21 |> 
  mutate(LATITUDE = if_else(LATITUDE %in% c(99.9999, 88.8888, 77.7777), NA, LATITUDE),
         LONGITUD = if_else(LONGITUD %in% c(999.9999, 888.8888, 777.7777), NA, LONGITUD))

subfolder_years <- c(2006:2015, 2020:2021)
no_subfolder_years <- c(2000:2005, 2016:2019)


person_no_subfolder <- map2_dfr(no_subfolder_years, 
                                  "person", 
                                  read_FARS_no_subfolder)
person_subfolder <- map2_dfr(subfolder_years, 
                               "person", 
                               read_FARS_subfolder)

person <- bind_rows(person_subfolder, person_no_subfolder)
accident <- bind_rows(acciden)

saveRDS(person, file = "data/person.RDS")
saveRDS(accident, file = "data/accident.RDS")

person <- readRDS("data/person.RDS")
accident2 <- readRDS("data/accident.RDS")
```

# Filtering and cleaning the data
FARS data include all traffic fatalities. We filter the person file to only include cases where the person is a pedestrian and the injury severity is a fatality. 

```{r}
peds <- person |> 
  filter(PER_TYP == 5 & INJ_SEV == 4)

View(x)
```

The crash locations are in the `ACCIDENT` table. Not all crashes are geocoded and we remove the ones that aren't.

```{r}
accident <- accident |> 
  filter(LONGITUD < 180 | LATITUDE < 70)
```

Now we can join the two tables, using the `ST_CASE` identifier.

```{r}
peds <- peds |> 
  left_join(accident, by = join_by(ST_CASE, 
                                   MONTH, 
                                   DAY, 
                                   HOUR, 
                                   MINUTE,
                                   STATE,
                                   VE_FORMS,
                                   COUNTY,
                                   RUR_URB,
                                   FUNC_SYS,
                                   HARM_EV,
                                   MAN_COLL,
                                   SCH_BUS))
```

The data contain a lot of variables and for now we'll only keep the ones needed for the time of data analysis, speeding up the computations.

```{r}
variables_for_daylight <- c(
  "STATE",
  "ST_CASE",
  "COUNTY",
  "DAY",
  "MONTH",
  "HOUR",
  "MINUTE",
  "YEAR",
  "AGE",
  "SEX",
  "HISPANIC",
  "RACE",
  "LATITUDE",
  "LONGITUD"
)

peds <- peds |> 
  select(all_of(variables_for_daylight))
```

We filter for crashes that don't contain an exact time or no location. We also parse the date and time fields.

```{r}
peds <- peds |> 
  filter(MINUTE != 99 & !is.na(LATITUDE)) |> #remove crashes with missing time, coords
  mutate(
    crash_time = ymd_hm(paste(YEAR, MONTH, DAY, HOUR, MINUTE, sep = "-")),
    crash_date = ymd(paste(YEAR, MONTH, DAY, sep = "-"))) 
```

Next we need to deal with time and time zones. The crash time and date in FARS appears to be stored as the local time, with no time zone information. We will use the latitude and longitude to look up the time zone and then convert all times into UTC.

```{r}
# create sf object
peds <- st_as_sf(peds, coords = c("LONGITUD", "LATITUDE"), remove = FALSE) #keep lat/lon columns    
peds$tz <- tz_lookup(peds, method = "accurate")

# Function to convert times to UTC
convert_to_utc <- function(datetime, timezone) {
  # Set the time zone of the datetime object
  datetime <- force_tz(datetime, tzone = timezone) #assign time zone
  
  # Convert the datetime to UTC
  datetime_utc <- with_tz(datetime, tzone = "UTC") #convert to time zone
  
  return(as_datetime(datetime_utc))
}

peds$crash_time_utc <- mapply(convert_to_utc, datetime = peds$crash_time, timezone = peds$tz)

peds <- peds |> 
  mutate(crash_time_utc = as_datetime(crash_time_utc),
         crash_date_utc = date(crash_time_utc))

```

Based on the new UTC times we can now calculate sunrise and sunsets for all crashes. This was more complicated than I thought: Once we get to calculating the time to/from the nearest sunrise/sunset, sometimes that sunrise or sunset falls to a different date. So we calculate the sunrise/sunset for the day before the crash, the day of the crash, and the day after, and then use the minimum difference. (I feel there must be a simpler way to do this...)

```{r}
peds <-  peds |> 
  rowwise() |> 
  mutate(sun_0 = getSunlightTimes(date = crash_date, lat = LATITUDE, lon = LONGITUD),
         sun_plus_1 = getSunlightTimes(date = crash_date + 1, lat = LATITUDE, lon = LONGITUD),
         sun_minus_1 = getSunlightTimes(date = crash_date - 1, lat = LATITUDE, lon = LONGITUD),
         sunrise_0 = sun_0$sunrise,
         sunrise_plus_1 = sun_plus_1$sunrise,
         sunrise_minus_1 = sun_minus_1$sunrise,
         sunset_0 = sun_0$sunset,
         sunset_plus_1 = sun_plus_1$sunset,
         sunset_minus_1 = sun_minus_1$sunset)

# saveRDS(peds, file = "data/peds_with_sunset.rds")
# peds <- readRDS("data/peds_with_sunset.rds")


peds <- peds |> 
  mutate(time_from_sunrise_0 = crash_time_utc - sunrise_0,
         time_from_sunset_0 = crash_time_utc - sunset_0,
         time_from_sunrise_minus_1 = crash_time_utc - sunrise_minus_1,
         time_from_sunset_minus_1 = crash_time_utc - sunset_minus_1,
         time_from_sunrise_plus_1 = crash_time_utc - sunrise_plus_1,
         time_from_sunset_plus_1 = crash_time_utc - sunset_plus_1,
         time_from_sunrise_min = case_when(
           min(abs(time_from_sunrise_0), abs(time_from_sunrise_minus_1), abs(time_from_sunrise_plus_1)) == abs(time_from_sunrise_0) ~ time_from_sunrise_0, 
           min(abs(time_from_sunrise_0), abs(time_from_sunrise_minus_1), abs(time_from_sunrise_plus_1)) == abs(time_from_sunrise_minus_1) ~ time_from_sunrise_minus_1,
           min(abs(time_from_sunrise_0), abs(time_from_sunrise_minus_1), abs(time_from_sunrise_plus_1)) == abs(time_from_sunrise_plus_1) ~ time_from_sunrise_plus_1),
         time_from_sunset_min = case_when(
           min(abs(time_from_sunset_0), abs(time_from_sunset_minus_1), abs(time_from_sunset_plus_1)) == abs(time_from_sunset_0) ~ time_from_sunset_0, 
           min(abs(time_from_sunset_0), abs(time_from_sunset_minus_1), abs(time_from_sunset_plus_1)) == abs(time_from_sunset_minus_1) ~ time_from_sunset_minus_1,
           min(abs(time_from_sunset_0), abs(time_from_sunset_minus_1), abs(time_from_sunset_plus_1)) == abs(time_from_sunset_plus_1) ~ time_from_sunset_plus_1)
         )

peds |> 
  ggplot(aes(as.numeric(time_from_sunset_min, units = "hours"))) +
  geom_histogram(binwidth = .25) +
  geom_vline(xintercept = 0, color = "red", alpha = 0.3) +
  facet_wrap(~YEAR) +
  ylab("Pedestrian fatalities") +
  xlab("Hours from sunrise") +
  labs(title = "Crashes are highly concentrated right after sunset",
       subtitle = "Fatal pedestrian crashes in the US, 2000-2021; 15-minute bins",
       caption = "Data: Fatality Analysis Reporting System (FARS)\nVisualization: Harald Kliems") +
  theme_ipsum(base_family = "Roboto Condensed")

x <- peds |> 
  st_drop_geometry() |> 
  mutate(minutes_since_midnight = HOUR*60 + MINUTE,
         int_15_min = cut(minutes_since_midnight, breaks = seq(1, 1440, by = 15)),
         int_
    )

x |> 
  group_by(int_15_min, MONTH) |> 
  summarize(n = n()) |> 
  group_by(MONTH) |> 
  mutate(highlight = ifelse(n == max(n), "yes", "no")) |> 
  ggplot(aes(int_15_min, n)) +
  geom_point() +
  facet_wrap(~MONTH, ncol = 1) +
  gghighlight::gghighlight(highlight == "yes", calculate_per_facet = TRUE,use_direct_label = FALSE) +
  theme_minimal()
  
peds |> 
  # left_join(state_fips, by = join_by(STATE.x == state_code)) |> 
  ggplot(aes(as.numeric(time_from_sunset_min, units = "hours"))) +
  geom_histogram(binwidth = .25) +
  geom_vline(xintercept = 0, color = "red", alpha = 0.3) +
  # facet_wrap(~YEAR) +
  ylab("Pedestrian fatalities") +
  xlab("Hours from sunset") +
  labs(title = "Crashes are highly concentrated right after sunset",
       subtitle = "Fatal pedestrian crashes in the US, 2016-2021; 15-minute bins",
       caption = "Data: Fatality Analysis Reporting System (FARS)\nVisualization: Harald Kliems") +
  theme_ipsum(base_family = "Roboto Condensed")

  ```



# Sunrise and sunset data
The time and date of the crashes appear to be saved in the local time zone where they occurred. To calculate sunrise and sunset data, we first convert the data to the UTC time zone. 



